Omics-based Single Step Trait Prediction

### 2017-09-12 paper/manuscript.tex
paper, cosmetic changes

\Genetics2level{Exp2}
To address objectives 1 and 2, a maize hybrid data set, subsequently
denoted as \textit{Exp2}, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines~\cite{Westhues2017}. 
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse locations across Germany for six 
agronomically important traits by accounting for year, location, field 
replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
using the Beagle software~\cite{Browning2009}.
The final number of polymorphic marker loci was 7,013 for the Dent and 6,212 for 
the Flint lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 Dent and 43 Flint lines, that are parents of 685 hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts while adjusting
for putative effects arising from the presence of multiple factorials
(\textit{cf.} \citeNP{Westhues2017}) using the R-package
\textit{limma}~\cite{Ritchie2015a}.

% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
\begin{table}[]
\centering
\caption{Acronyms for the trait names}
\label{table:TraitKey}
\begin{tabular}{@{}lll@{}}
\toprule
Trait              & Abbreviation & Data Set  \\ \midrule
Cob weight         & CW           & Diversity \\
Days to silking    & DS           & Diversity \\
Ear diameter       & ED           & Diversity \\
100-grain weight   & GW           & Diversity \\
Kernel width       & KW           & Diversity \\
Plant height       & PH           & Diversity \\
Dry matter yield   & DMY          & Hybrid    \\
Dry matter content & DMC          & Hybrid    \\
Fat                & FAT          & Hybrid    \\
Protein            & PRO          & Hybrid    \\
Starch             & STA          & Hybrid    \\
Sugar              & SUG          & Hybrid    \\ \bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/inbred_pca.pdf}
  \caption{
  Principal component analysis of 211 genotypes from the set
  `tropical/subtropical' of the maize diversity panel inbred lines.
  The four clusters are based on $K=4$ putative ancestral populations in a
  STRUCTURE analysis of the corresponding genomic feature matrix.
  Genotypes are colored based on their primary assignment to any of these four
  clusters.
}
\label{fig:inbred-pca}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=1\textwidth]{./tables_figures/compound_plot.pdf}
  \caption{
  (\textbf{a}) Scheme for the amount of genotypes covered by predictors.
  The two extended bars represent the entire set of genotypes in either
  \textit{Exp2} or \textit{Exp1}.
  All genotypes are covered by at least one predictor, which is called the
  `complete' predictor and depicted as a white bar.
  The second bar depicts that only a fraction of genotypes is covered by both,
  the `complete' and the `incomplete' predictor (dark gray) and that the
  difference between the predictors will be imputed.
  The four bars to the right outline the principle of generating core sets for
  \textit{Exp1}.
  For nine different values of $x$, which represents the fraction of genotypes
  covered by the `incomplete' predictor, core sets are build.
  (\textbf{b}) Subsampling within core sets.
  Each of the black symbols represents a different genotype.
  For each value of $x$, 100 subsets $s$ are generated where $x\%$ of genotypes
  are randomly declared to be covered by both, the `complete' and the
  `incomplete' predictor whereas the remaining $100\% - x\%$ of genotypes
  declared to be covered only by the `complete' predictor.
  }
\label{fig:compound-plot}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/pred_ability_inbred.pdf}
  \caption{
  Predictive abilities and SE for tropical/subtropical inbred lines from the 
  maize diversity panel and six agronomic traits.
  As predictors, genomic (G) data, transcriptomic (T) data and their 
  combination (GT) were used.
  The `Reduced' panel includes a subset of 110 lines covered by genomic and
  transcriptomic information whereas the `Full' panel comprises all 164
  genotypes.
  In the `Full' set of genotypes, transcriptomic records were imputed for 54
  lines.
  }
\label{fig:InbredResults}
\end{figure}

\Genetics2level{Exp2}
The maize hybrid data set contains information on seven agronomic traits and
three predictors, namely pedigree data (P), genomic information (G), 
transcriptomic information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the `Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as `Reduced'.
In the `core' set, transcriptomic data were the best single predictor for ADL,
PRO and particulaly for DMY where the predictive ability obtained with T
was 0.08 points higher than for G as the second best predictor for this trait
(Fig.~\ref{fig:HybridResults}).
In the core set, P had the highest predictive ability for STA and DMC\@.
For the FAT and SUG, G was the superior predictor inside the core set.
In the full set of 1,521 hybrids, predictive abilities obtained with G rose 
markedly for all traits with the exception of FAT\@.
The combination of genomic and transriptomic information (GT) in a single-step
model yielded a small improvement over G alone for the traits DMY and PRO\@.
Except for PRO, compared to using only G, trait prediction did not benefit from 
the single step approach when T was not superior to G in the core set.
Predictive abilities resulting from combinations of pedigree with genomic (PG)
information were better than P alone for all traits and PT improved upon P in
the `full' set for four traits.

\begin{figure}[H]
\centering
  \includegraphics{./tables_figures/pred_ability_hybrid.pdf}
  \caption{
    Predictive abilities and average, bootstrapped standard errors for the set
    of maize hybrids and six agronomic traits.
    As predictors, pedigree (P), genomic (G), transcriptomic (T) data and 
    combinations thereof we used.
    The `Full' set of genotypes comprises 1,521 hybrid progeny based on 245
    parental inbred lines, whereas the `Reduced' set comprises 685 hybrid
    progeny based on 103 parental inbred lines.
  }
\label{fig:HybridResults}
\end{figure}

@@

Predictive abilities obtained with the different core sets varied widely within
most traits (Fig.~\ref{fig:InbredResults}).
For most traits, an initial increase in predictive abilities when moving
from a core set size of 0.1 to 0.2/0.3 was observed, regardless of whether G
or T individually was the superior predictor.
For the traits CW, GW and PH, which were better predicted via G than via T in
the reduced set of lines, predictive abilities were higher when only a small
number of genotypes was included in the core set.
This corresponds to a situation where a large number of genotypes is being 
imputed and the majority of the information is derived from genomic information.
A hike in predictive abilities, when moving from a core fraction of 10\% to 
90\%, was observed for ED and KW, which were predicted better when using T
individually compared to using G individually.
The only trait that was barely affected by a change in the core fraction was DS
for which the predictive abilities using either G or T individually were almost
identical (Fig.~\ref{fig:InbredResults}).

@@

\Genetics2level{Superiority of single step models}
\subsection{Exp1}
When using a single step approach based on complete genomic and incomplete
transcriptomic information, gains in predictive ability for the inbred lines
were at best just slightly higher than when using complete genomic information
alone.
One of the reasons might be that the subset of genotypes that was covered by
transcriptomic data was not a good representation of the full genetic space of
all available genotypes.
Another possibility is that the number of additional phenotypes in the full
set compared to the core set ($\Delta(n) = 54$) was not large enough to provide
considerably more useful information.
The latter was the case for \citeNP{Ashraf2016}, who imputed genotypes for about 
10,000 wheat lines using pedigree information and observed greater accuracy of
the single step method over genomic BLUP for all four evaluated traits.
Predictive abilities reported for the maize diversity panel in this study were
slightly different from those reported by \citeNP{Guo2016}, for three
possible reasons:
1) We reduced the dataset to 164 tropical/subtropical lines with little evident
population structure and excluded all other genotypes from our analyses whereas
\citeNP{Guo2016} used genotypes from four different subpopulations, modeled as a
fixed effect.
2) We applied quality checks for the predictor data after generating the subset
of 164 inbred lines, whereas \citeNP{Guo2016} applied the quality checks to 368
genotypes.
3) While \citeNP{Guo2016} used five-fold cross-validation with 500 repetitions,
we employed a leave-one-out cross-validation scheme (LOOCV).
Notwithstanding, the relative differences between predictive abilities for G 
and T, respectively, were similar in both studies.

\subsection{Exp2}
Hybrid breeding --- with noteworthy commercial applications in pig
\cite{Xiang2016,Tusell2016} and maize breeding~\cite{TheRoyalSociety2009} --- 
is a particularly challenging application field for prediction tools 
\cite{Kadam2016}.
Here, $2n$ parent individuals from two genetically distinct heterotic groups are
crossed to each other, yielding $n^{2}$ potential hybrid progeny that would
require intensive field testing.
In medium-sized plant breeding programs, the advent of the doubled haploid (DH) 
technology~\cite{Wedzony2009} now allows for an annual production of 1,000 
parent lines in each heterotic group, amounting to $10^{6}$ putative hybrid
progeny.
\citeNP{Westhues2017} have shown that, in hybrid breeding, the probability 
of successfully selecting the best observed genotypes based on the best 
predicted candidates is a strongly convex function of the predictive ability.
Thus, even the minor gains when using transcriptomic and genomic data in a
single step might justify additional investments in RNA-seq at least for a
subset of genotypes that represents the genetic space of the breeding 
material well.
For the hybrid data, the predictive abilities of genomic information
for the whole set of available parent lines rose considerably for several
agronomic traits.
Hence, the inclusion of trancsriptomic data could not further improve upon
this result.
Predictive abilities obtained when using pedigree data could be 
improved considerably when, depending on the agronomic trait, combined with 
either transcriptomic or genomic information.

\Genetics2level{Coverage of the genetic space}
Whereas animal breeding populations are oftentimes so large that the assembly 
of a core set of individuals with records on multiple predictors can effectively 
be done at random once more than 10,000 animals have pedigree records 
\cite{Fragomeni2015,Lourenco2015,Masuda2016}, pedigrees in plant breeding 
programs are typically considerably smaller.
Single step prediction offers the promise of leveraging the predictive ability
for all genotypes by borrowing the superior performance of an auxiliary
predictor on only a subset of genotypes, thereby reducing predictor costs.
Hence, we were interested in determining which individuals should be 
complemented with information on another predictor.
By randomly and repeatedly declaring trancsriptomic information missing from
genotypes in \textit{Exp1} we examined how reliably single step prediction works
depending on the fraction of genotypes covered by both, the complete and the
incomplete predictor.
We observed that about $20--30\%$ of genotypes needed to be covered by both
predictors to match the performance achieved with the complete predictor
individually.
Beyond this fraction of covered genotypes the trend in predictive ability
changes was stable.

\section*{Conclusions}
We have successfully applied single step prediction to a hybrid plant data set 
while imputing with a quantitative downstream `omics' predictor.
By declaring different subsets of individuals as covered by one or two 
predictors, respectively, we could elucidate the influence of differences
between two predictors on predictive ability when used in a single step
approach.
When the discrepancy in predictive ability between the complete and the
incomplete predictor was skewed profoundly in favor of the incomplete predictor,
such as for pedigree and transcriptomic data, single step prediction was shown
be highly successful.
Extensions to more than two predictors were outside the scope of this study
but, with mounting interest in systems genetics, should be considered in the
future.

>>>>

\Genetics2level{Exp2}
To address objectives 1 and 2, a maize hybrid data set, subsequently
denoted as \textit{Exp2}, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines~\cite{Westhues2017}. 
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse locations across Germany for seven
agronomically important traits by accounting for year, location, field 
replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
using the Beagle software~\cite{Browning2009}.
The final number of polymorphic marker loci was 7,013 for the Dent and 6,212 for 
the Flint lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 Dent and 43 Flint lines, that are parents of 685 hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts while adjusting
for putative effects arising from the presence of multiple factorials
(\textit{cf.} \citeNP{Westhues2017}) using the R-package
\textit{limma}~\cite{Ritchie2015a}.

% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
\begin{table}[]
\centering
\caption{Acronyms for the trait names}
\label{table:TraitKey}
\begin{tabular}{@{}lll@{}}
\toprule
Trait              & Abbreviation & Data Set  \\ \midrule
Cob weight         & CW           & Diversity \\
Days to silking    & DS           & Diversity \\
Ear diameter       & ED           & Diversity \\
100-grain weight   & GW           & Diversity \\
Kernel width       & KW           & Diversity \\
Plant height       & PH           & Diversity \\
Dry matter yield   & DMY          & Hybrid    \\
Dry matter content & DMC          & Hybrid    \\
Lignin             & ADL          & Hybrid    \\
Fat                & FAT          & Hybrid    \\
Protein            & PRO          & Hybrid    \\
Starch             & STA          & Hybrid    \\
Sugar              & SUG          & Hybrid    \\ \bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/inbred_pca.pdf}
  \caption{
  Principal component analysis of 211 genotypes from the set
  `tropical/subtropical' of the maize diversity panel inbred lines.
  The four clusters are based on $K=4$ putative ancestral populations from a
  STRUCTURE analysis of the corresponding genomic feature matrix.
  Genotypes are colored based on their primary assignment to any of these four
  clusters.
}
\label{fig:inbred-pca}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=1\textwidth]{./tables_figures/compound_plot.pdf}
  \caption{
  (\textbf{a}) Scheme for the amount of genotypes covered by predictors.
  The two extended bars represent the entire set of genotypes in either
  \textit{Exp2} or \textit{Exp1}.
  All genotypes are covered by at least one predictor, which is called the
  `complete' predictor and depicted as a white bar.
  The second bar shows that only a fraction of genotypes is covered by both,
  the `complete' and the `incomplete' predictor (dark gray) and that the
  difference between the predictors will be imputed (light gray fraction).
  The four bars to the right outline the principle of generating core sets for
  \textit{Exp1}.
  For nine different values of $x$, which represents the fraction of genotypes
  covered by the `incomplete' predictor, core sets are build.
  (\textbf{b}) Subsampling within core sets.
  Each of the black symbols represents a different genotype.
  For each value of $x$, 100 subsets $s$ are generated where $x\%$ of genotypes
  are randomly declared to be covered by both, the `complete' and the
  `incomplete' predictor whereas the remaining $100\% - x\%$ of genotypes
  are declared to be covered only by the `complete' predictor.
  }
\label{fig:compound-plot}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/pred_ability_inbred.pdf}
  \caption{
  Predictive abilities and average, bootstrapped standard errors for
  tropical/subtropical inbred lines from the maize diversity panel and six
  agronomic traits.
  As predictors, genomic (G) data, transcriptomic (T) data and their 
  combination (GT) were used.
  The `Reduced' panel includes a subset of 110 lines covered by genomic and
  transcriptomic information whereas the `Full' panel comprises all 164
  genotypes.
  In the `Full' set of genotypes, transcriptomic records were imputed for 54
  lines.
  }
\label{fig:InbredResults}
\end{figure}

\Genetics2level{Exp2}
The maize hybrid data set contains information on seven agronomic traits and
three predictors, namely pedigree data (P), genomic information (G), 
transcriptomic information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the `Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as `Reduced'.
In the `Reduced' set, transcriptomic data were the best single predictor for
ADL, PRO and particulaly for DMY where the predictive ability obtained with T
was 0.08 points higher than for G as the second best predictor for this trait
(Fig.~\ref{fig:HybridResults}).
In the reduced set, P had the highest predictive ability for STA and DMC\@.
For FAT and SUG, G was the superior predictor inside the reduced set.
In the full set of 1,521 hybrids, predictive abilities obtained with G rose 
markedly for all traits with the exception of FAT\@.
The combination of genomic and transriptomic information (GT) in a single-step
model yielded a small improvement over G alone for the traits DMY and PRO\@.
Except for PRO, compared to using only G, trait prediction did not benefit from 
the single step approach when T was not superior to G in the reduced set.
Predictive abilities resulting from combinations of pedigree with genomic (PG)
information were better than P alone for all traits and PT improved upon P in
the `full' set for four traits.

\begin{figure}[H]
\centering
  \includegraphics{./tables_figures/pred_ability_hybrid.pdf}
  \caption{
    Predictive abilities and average, bootstrapped standard errors for the set
    of maize hybrids and seven agronomic traits.
    As predictors, pedigree (P), genomic (G), transcriptomic (T) data and 
    combinations thereof we used.
    The `Full' set of genotypes comprises 1,521 hybrid progeny based on 245
    parental inbred lines, whereas the `Reduced' set comprises 685 hybrid
    progeny based on 103 parental inbred lines.
  }
\label{fig:HybridResults}
\end{figure}

@@

Predictive abilities obtained with the different core sets varied widely within
most traits (Fig.~\ref{fig:InbredResults}).
For most traits, an initial increase in predictive abilities when moving
from a core set size of 10\% to 20\%/30\% was observed, regardless of whether G
or T individually was the superior predictor.
For the traits CW, GW and PH, which were better predicted via G than via T in
the reduced set of lines, predictive abilities were higher when only a small
number of genotypes was included in the core set.
This corresponds to a situation where a large number of genotypes is being 
imputed and the majority of the information is derived from genomic information.
A hike in predictive abilities, when moving from a core fraction of 10\% to 
90\%, was observed for ED and KW, which were predicted better when using T
individually compared to using G individually.
The only trait that was barely affected by a change in the core fraction was DS
for which the predictive abilities using either G or T individually were almost
identical (Fig.~\ref{fig:InbredResults}).

@@

\Genetics2level{Superiority of single step models}
\subsection{Exp1}
When using a single step approach based on complete genomic and incomplete
transcriptomic information, gains in predictive ability for the inbred lines
were at best just slightly higher than when using complete genomic information
alone.
One of the reasons might be that the subset of genotypes that was covered by
transcriptomic data was not a good representation of the full genetic space of
all available genotypes.
Another possibility is that the number of additional phenotypes in the full
set compared to the reduced set ($\Delta(n) = 54$) was not large enough to
provide considerably more useful information.
The latter was the case for \citeNP{Ashraf2016}, who imputed genotypes for about 
10,000 wheat lines using pedigree information and observed greater accuracy of
the single step method over genomic BLUP for all four evaluated traits.
Predictive abilities reported for the maize diversity panel in this study were
slightly different from those reported by \citeNP{Guo2016}, for three
possible reasons:
1) We reduced the dataset to 164 tropical/subtropical lines with little evident
population structure and excluded all other genotypes from our analyses whereas
\citeNP{Guo2016} used genotypes from four different subpopulations, modeled as a
fixed effect.
2) We applied quality checks for the predictor data after generating the subset
of 164 inbred lines, whereas \citeNP{Guo2016} applied the quality checks to 368
genotypes.
3) While \citeNP{Guo2016} used five-fold cross-validation with 500 repetitions,
we employed a leave-one-out cross-validation scheme (LOOCV).
Notwithstanding, the relative differences between predictive abilities for G 
and T, respectively, were similar in both studies.

\subsection{Exp2}
Hybrid breeding --- with noteworthy commercial applications in pig
\cite{Xiang2016,Tusell2016} and maize breeding~\cite{TheRoyalSociety2009} --- 
is a particularly challenging application field for prediction tools 
\cite{Kadam2016}.
Here, $2n$ parent individuals from two genetically distinct heterotic groups are
crossed to each other, yielding $n^{2}$ potential hybrid progeny that would
require intensive field testing.
In medium-sized plant breeding programs, the advent of the doubled haploid (DH) 
technology~\cite{Wedzony2009} now allows for an annual production of 1,000 
parent lines in each heterotic group, amounting to $10^{6}$ putative hybrid
progeny.
\citeNP{Westhues2017} have shown that, in hybrid breeding, the probability 
of successfully selecting the best observed genotypes based on the best 
predicted candidates is a strongly convex function of the predictive ability.
Thus, even minor gains, when using transcriptomic and genomic data in a
single step, might justify additional investments in RNA-seq at least for a
subset of genotypes that represents the genetic space of the breeding 
material well.
For the hybrid data, the predictive abilities of genomic information
for the whole set of available parent lines rose considerably for several
agronomic traits.
Hence, the inclusion of trancsriptomic data could not further improve upon
genomic information alone.
Predictive abilities obtained when using pedigree data could be 
improved considerably when, depending on the agronomic trait, combined with 
either transcriptomic or genomic information.
This suggests that single step prediction should always be considered when the
incomplete predictor offers an appreciably higher predictive ability compared
to the complete predictor.

\Genetics2level{Coverage of the genetic space}
Whereas animal breeding populations are oftentimes so large that the assembly 
of a core set of individuals with records on multiple predictors can effectively 
be done at random once more than 10,000 animals have pedigree records 
\cite{Fragomeni2015,Lourenco2015,Masuda2016}, pedigrees in plant breeding 
programs are typically considerably smaller.
Single step prediction offers the promise of leveraging the predictive ability
for all genotypes by borrowing the superior performance of an auxiliary
predictor on only a subset of genotypes, thereby reducing predictor costs.
Hence, we were interested in determining to what extent individuals should be 
complemented with information on another predictor.
By randomly and repeatedly declaring trancsriptomic information missing from
genotypes in \textit{Exp1} we examined how reliably single step prediction works
depending on the fraction of genotypes covered by both, the complete and the
incomplete predictor.
We observed that about $20-30\%$ of genotypes needed to be covered by both
predictors to match the performance achieved with the complete predictor
individually.
Beyond this fraction of covered genotypes the trend in predictive ability
changes was stable.

\section*{Conclusions}
We successfully applied single step prediction to a hybrid plant data set 
while imputing with a quantitative downstream `omics' predictor.
By declaring different subsets of individuals as covered by one or two 
predictors, respectively, we could elucidate the influence of differences
between two predictors on predictive ability when used in a single step
approach.
When the discrepancy in predictive ability between the complete and the
incomplete predictor was skewed profoundly in favor of the incomplete predictor,
such as for pedigree and transcriptomic data, single step prediction was shown
to be highly beneficial.
Extensions to more than two predictors were outside the scope of this study
but, with mounting interest in systems genetics, should be considered in the
future.
